Optimizing the future of biodiversity sampling by citizen scientists

### 2019-04-09 paper/main.tex
bit more crap on paper

\title{Optimizing biodiversity sampling by citizen scientists}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{A}ssessing biodiversity trends in space and time is essential for conservation \cite{harrison2014assessing, wilson2011modelling, mcmahon2011improving, honrado2016fostering} \parshape=0, allowing the tracking of progress, and assessment of whether biodiversity targets are being met. Unsurprisingly, trend detection is best estimated when using data from well-designed surveys in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}, that are long-term \cite{lindenmayer2012value}. But scientific funding for long-term ecological research is rapidly disappearing [REFS???]. Increasingly, government agencies, scientific researchers, and conservationists are turning to citizen science data to help inform the state of biodiversity at local \cite{callaghan2015efficacy, theobald2015global, sullivan2017using, loss2015linking}, regional \cite{barlow2015citizen, fox2011new}, and global scales \cite{chandler2017contribution, pocock2018vision, cooper2014invisible}.

Citizen science projects fall along a continuum of the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., iNaturalist) to structured (U.K. Butterfly Monitoring Scheme). All citizen science projects have associated biases, including spatial and temporal biases \cite{boakes2010distorted}, and different levels of observers' skills \cite{kelling2015can}. But these biases can generally be accounted for, statistically \cite{isaac2014statistics, robinson2018correcting}, by filtering or subsetting the data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}. Regardless of biases, estimating trends is best done with structured projects \cite{fox2011new}, but unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, @kery2010site, @horns2018using, @van2013occupancy, @pagel2014quantifying}.

The number of ecological and environmental monitoring citizen science projects is increasing \cite{pocock2017diversity, theobald2015global}, highlighting the potential of these data in the future of ecology, conservation, and resource management \cite{pocock2018vision, mckinley2017citizen}. But a major obstacle in the future use of citizen science data remains understanding how to best extract information from 'noisy' citizen science datasets \cite{parrish2018exposing}.

Clearly, citizen science holds great potential for the future of ecology, conservation, and natural resource management \cite{pocock2018vision, silvertown2009new, soroye2018opportunistic, mckinley2017citizen}. But can it be improved? Are there mechanisms which can help diminish the spatial and temporal biases that exist in citizen science projects and associated data?

Here, we investigate these questions, and our objective is to optimize citizen science sampling for trend detection of biodiversity. Many citizen science projects have specific biases and quality concerns. For instance, observer skill \cite{kelling2015can}, time-of-day, number of participants in a group, and technological capabilities of a participant may influence the data collected by some, but not all, citizen science projects. However, generalized across all citizen science projects are spatial and temporal biases \cite{boakes2010distorted}. For instance, biases such as intra-week differences in sampling \cite{courter2013weekend}, sampling near roads and high-density population centers \cite{kelling2015taking}, and XXXX are biases that are generalizable across citizen science projects.

We predicted that the following parameters would be positively associated with the marginal value of a given citizen science observation: whether a site was sampled or not, median sampling interval, days since last sample, distance to the nearest sampled site, nearest neighbor sampling interval, and the number of unique days sampled.

\begin{itemize}
  \item \textbf{Whether a site was sampled}: if a site had been previously sampled or not.
  \item \textbf{Median sampling interval}: the median of the distribution of waiting times between samples at a site.
  \item \textbf{Days since last sample}: the number of days between samples at a site.
  \item \textbf{Distance to the nearest sampled site}: the distance between the site in question and the nearest sampled site.
  \item \textbf{Nearest neighbor sampling interval}: the median sampling interval of the nearest neighbor.
  \item \textbf{Number of unique days sampled}: the total number of unique days sampled for a given site.
\end{itemize}

Here, we illustrate our vision using a popular citizen science project --- eBird \cite{sullivan2009ebird}.

\section*{Results}

\section*{Discussion}
Citizen science projects need to continuously adapt to be successful, and to better inform conservation science. Even if a project may not have specific objectives/goals, it is important for the project to adapt to particular questions.

@@

>>>>

\title{Optimizing the future of biodiversity sampling by citizen scientists}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{A}ssessing biodiversity trends in space and time is essential for conservation \cite{harrison2014assessing, wilson2011modelling, mcmahon2011improving, honrado2016fostering} \parshape=0. Reliable biodiversity estimates allow us to track our global progress in curbing biodiversity loss \cite{harrison2014assessing}. Unsurprisingly, reliable trend detection is usually derived from well-designed surveys in space and time \cite{harrison2014assessing, vellend2017estimates, kery2009trend}, that are long-term \cite{lindenmayer2012value, magurran2010long}. But scientific funding for long-term ecological and conservation research is rapidly disappearing [REFS???]. Increasingly, government agencies, scientific researchers, and conservationists are turning to citizen science data to help inform the state of biodiversity at local \cite{callaghan2015efficacy, theobald2015global, sullivan2017using, loss2015linking}, regional \cite{barlow2015citizen, fox2011new}, and global scales \cite{chandler2017contribution, pocock2018vision, cooper2014invisible}.

Citizen science --- the cooperation between a range of experts and non-experts  --- is an incredibly diverse field \cite{jordan2015citizen}. Projects generally fall along a continuum based on the level of associated structure \cite{kelling2019using, welvaert2016citizen}, ranging from unstructured (e.g., iNaturalist) to structured (U.K. Butterfly Monitoring Scheme). The level of structure, in turn, influences the biases and data quality concerns of a particular project. For instance, observer skill \cite{kelling2015can}, time-of-day, number of participants in a group, and technological capabilities of a participant may influence the data collected by some, but not necessarily all, citizen science projects. Generalizeable among citizen science projects, however, are various spatial and temporal biases \cite{boakes2010distorted, bird2014statistical}. Observers submitting observations on weekends \cite{courter2013weekend}, sampling near roads and human settlements \cite{kelling2015taking}, and observers visiting known 'hotspots' for biodiversity are all examples of spatial and temporal biases. Importantly, these biases are not restricted to citizen science projects. 'Professional scientists' also have biased sampling of biodiversity, highlighted by biased museum collections [REFS???].

Fortunately, in the case of broad-scale biodiversity data collected by citizen scientists, these various biases can generally be accounted for, statistically \cite{isaac2014statistics, robinson2018correcting}, by filtering or subsetting the data \cite{wiggins2011conservation}, pooling multiple data sources \cite{fithian2015bias}, or machine learning and hierarchical clustering techniques \cite{hochachka2012data, kelling2015taking}. Despite these biases, citizen science data have increased our knowledge on species distribution models \cite{bradsworth2017species, van2013opportunistic}, niche breadth \cite{tiago2017using}, biodiversity measurements \cite{stuart2017assessing, pocock2018vision}, phenological research \cite{la2014role, supp2015citizen}, invasive species detection \cite{pocock2017citizen, grason2018citizen}, and phylogeographical research \cite{bahls2014new, drury2019continent}. Still, estimating trends with citizen science data is best done with data from structured projects (i.e., less biases to account for) \cite{fox2011new}, but unstructured and semistructured projects are increasingly harnessed for trend detection \cite{walker2017using, kery2009trend, kery2010site, horns2018using, van2013occupancy, pagel2014quantifying}.

The number of ecological and environmental monitoring citizen science projects is increasing \cite{pocock2017diversity, theobald2015global}, highlighting the potential that citizen science holds for the future of ecology, conservation, and natural resource management \cite{pocock2018vision, silvertown2009new, soroye2018opportunistic, mckinley2017citizen}. But a major obstacle in the future use of citizen science data remains understanding how to best extract information from 'noisy' citizen science datasets \cite{parrish2018exposing}.

However, these 'big-data' statistical approaches \cite{kelling2015taking} are most applicable for data originating from large, successful citizen science projects. What about projects that are just starting? Or projects focusing on taxa that are less popular with the general public \cite{mair2016explaining, ward2014understanding}? Are there optimal strategies for sampling in space and time, which are best able to contribute to estimating biodiversity trends?

Here, we investigate these questions with a specific objective: investigate how spatial and temporal sampling influences trend detection of biodiversity. Our approach is site-specific and dynamic: we are interested in the parameters that influence the marginal value of a given citizen science observation for a particular day. Here, we summarize the parameters in space and time which we hyptohesized would influence the value of a citizen science observation.

\begin{itemize}
  \item \textbf{Whether a site was sampled}: if a site had been previously sampled or not. We predicted that unsampled sites would be marginally more valuable than sampled previously sampled sites.
  \item \textbf{Median sampling interval}: the median of the distribution of waiting times between samples at a site. We predicted that the median sampling interval would be positively associated with the value of a citizen science observation. I.e., observations from sites with high median waiting times would be more valuable than observations from sites with low median waiting times.
  \item \textbf{Days since last sample}: the number of days between samples at a site. We predicted that the days since the last sample would positively associate with the value of a citizen science observation.
  \item \textbf{Distance to the nearest sampled site}: the distance between the site in question and the nearest sampled site. We hypothesized that for the distance to the nearest sampled site would be positively associated with the value of a citizen science observation.
  \item \textbf{Nearest neighbor sampling interval}: the median sampling interval of the nearest neighbor. We hypothesized that this would be positively influence the value of a observation, whereby well-sampled areas (i.e., multiple sites near each other with low median sampling intervals) would have lower value observations.
  \item \textbf{Number of unique days sampled}: the total number of unique days sampled for a given site. We predicted that the total number of unique days would be positively associated with the value of an observation, whereby sites with lots of observations would receive value given the long-term data originating from them.
\end{itemize}

We illustrate our approach using a popular citizen science project --- eBird \cite{sullivan2009ebird} --- and data from the Greater Sydney Region, NSW, Australia.

\section*{Results and Discussion}
We found weak evidence that unsampled sites were more marginally valuable than unsampled sites across grid sizes (Fig. S1), noting, however, that there were no unsampled grids in the 50 km grid cell size. This suggests that stratified sampling --- an approach which aims for equal sampling among grids or sites --- [REF] may not be the most applicable approach for detecting trends using citizen science data. This also suggests that citizen scientists generally are sufficiently sampling the appropriate areas: local knowledge indicates where birders should go birding.

However, there are still parameters that are significantly important for the marginal value of a citizen science observation (PARAMETER ESTIMATES FIGURE). At all four grid sizes, days since the last sample was positively associated with the marginal value of a citizen science observation. For all but the 10 km grid size, distance to the nearest sampled grid was positively associated with the marginal value of a citizen science observation. Conversely, the results for median sampling interval were negatively associated with the value of a citizen science observation. Nearest neighbor sampling interval had relatively weak influence on the value of a citizen science observation, compared with other parameters, as did the number of unique days sampled for a given grid.

These results suggest that the value of a given citizen science observation can indeed be predicted based on temporal and spatial sampling. Through time, the longer a grid goes unsampled (i.e., days since last sample), that grid's marginal value will continue to increase. Similarly, this is synergistic with the distance to the nearest sampled grid, whereby the further away from a sampled grid an observation is, the more valuable that observation is.

We were also able to use our fitted models to predict the expected value of a given citizen science observation, dynamically, for any given day. This approach required us to look backward first, in order to predict future values. (STATIC MAP FIGURE).

Citizen science projects need to continuously adapt to be successful, and to better inform conservation science. Even if a project may not have specific objectives/goals, it is important for the project to adapt to particular questions.

@@

\subsection*{Data availability}
